---
title: "R Notebook"
output: html_notebook
---

# packages

```{r}

library(tidyverse)

```

# data

## race

```{r}

race <- c("no race", "white", "black") %>% rep(each = 5, times = 2) %>% factor()

```

## ecology

```{r}

ecology <- c("hopeful", "desperate") %>% rep(each = 15) %>% factor()

```

## group

```{r}

group <- interaction(race, ecology, sep = " and ")

```

## stereotype dv

```{r}

# set random seed so values reproduce
set.seed(1234)

stereotype <- c(rnorm(n = 5, mean = 1, sd = 1),
                rnorm(n = 5, mean = 0, sd = 1),
                rnorm(n = 5, mean = -1, sd = 1),
                rnorm(n = 5, mean = -1, sd = 1),
                rnorm(n = 5, mean = 0, sd = 1),
                rnorm(n = 5, mean = 1, sd = 1))

```

# plot means

```{r}

tibble(race, ecology, stereotype) %>% 
  ggplot(mapping = aes(x = ecology, y = stereotype, fill = race)) +
  stat_summary(geom = "bar", fun.data = mean_cl_normal, position = position_dodge(width = 0.9))

```

# dummy codes

## dummy code matrix

```{r}

tibble(race = c("black", "no race", "white"),
       race_d1 = c(0, 1, 0),
       race_d2 = c(0, 0, 1))
       
```

## race

```{r}

race_d1 <- ifelse(race == "black", 0,
           ifelse(race == "no race", 1,
           ifelse(race == "white", 0, NA)))

race_d2 <- ifelse(race == "black", 0,
           ifelse(race == "no race", 0,
           ifelse(race == "white", 1, NA)))

```

## ecology

```{r}

ecology_d1 <- ifelse(ecology == "desperate", 0,
              ifelse(ecology == "hopeful", 1, NA))

```

## interaction 1: race_d1 x ecology_d1

```{r}

rd1_ed1 <- race_d1 * ecology_d1

```

## interaction 2: race_d2 x ecology_d1

```{r}

rd2_ed1 <- race_d2 * ecology_d1

```

# dummy code model

## fit

```{r}

dummy_lm <- lm(stereotype ~ race_d1 + race_d2 + ecology_d1 + rd1_ed1 + rd2_ed1)

```

## summary

```{r}

dummy_lm %>% summary()

```

# effect codes
> center the dummy codes
> recompute interactions

```{r}

race_e1 <- race_d1 - mean(race_d1)
race_e2 <- race_d2 - mean(race_d2)
ecology_e1 <- ecology_d1 - mean(ecology_d1)
re1_ee1 <- race_e1 * ecology_e1
re2_ee1 <- race_e2 * ecology_e1

```

# effect code model

## fit

```{r}

effect_lm <- lm(stereotype ~ race_e1 + race_e2 + ecology_e1 + re1_ee1 + re2_ee1)

```

## summary

```{r}

effect_lm %>% summary()

```

# contrast codes

## race

```{r}

race_c1 <- ifelse(race == "black", -1,
           ifelse(race == "no race", 1,
           ifelse(race == "white", 0, NA)))

race_c2 <- ifelse(race == "black", -1,
           ifelse(race == "no race", 0,
           ifelse(race == "white", 1, NA)))

```

## ecology

```{r}

ecology_c1 <- ifelse(ecology == "desperate", -1,
              ifelse(ecology == "hopeful", 1, NA))

```

## interaction 1: race_c1 x ecology_c1

```{r}

rc1_ec1 <- race_c1 * ecology_c1

```

## interaction 2: race_c2 x ecology_c1

```{r}

rc2_ec1 <- race_c2 * ecology_c1

```

# contrast code model

## fit

```{r}

contrast_lm <- lm(stereotype ~ race_c1 + race_c2 + ecology_c1 + rc1_ec1 + rc2_ec1)

```

## summary

```{r}

contrast_lm %>% summary()

```



