---
title: 'Simple slopes in R and SPSS: Use the regression equation'
author: "Nicholas Michalak"
date: "10/14/2017"
output: html_document
---

## packages

```{r, message = FALSE, warning = FALSE}

library(tidyverse)
library(multcomp)
library(effects)

```

## set random seed

```{r}

set.seed(100)

```

## create the main predictor variable, x

```{r}

x <- rnorm(n = 100, mean = 0, sd = 1)

```

## create the continuous moderating variable, z

```{r}

z <- rnorm(n = 100, mean = 0, sd = 1) 

```

## create the categorical moderator, cat

```{r}

cat <- rep(c(-1, 1), each = 50)

```

## create the outcome variable, y

```{r}

y <- x * z * cat + rnorm(n = 100, mean = 0, sd = 1)

```

## store these in a data.frame

```{r}

example_data <- tibble(y, x, cat, z)

```

## create new variable that cuts x and z into quantiles

#### function for cutting into quantiles

```{r}

cut_quants <- function(x) {
  ## get quantiles
  qs <- quantile(x)
  
  ## cut x by those quantiles; output is factor
  cut(x, breaks = unique(qs), include.lowest = TRUE)
}

```

#### now make those variables

```{r}

example_data <- example_data %>% 
  mutate(x_quant = x %>% cut_quants(),
         z_quant = z %>% cut_quants())

```

## plot relationship, no regression lines

```{r}

example_data %>% 
  ggplot(mapping = aes(x = x, y = y, color = factor(cat))) +
  geom_point() +
  facet_wrap(facets = ~ z_quant)

```

## fit linear regression

```{r}

lm_01 <- lm(y ~ x * cat * z, data = example_data)

```

## summary of that regression

```{r}

lm_01 %>% summary()

```

## plot regression slopes
> first make sure there are actually z and z values across eachother's distribution

```{r}

example_data %>% 
  ggplot(mapping = aes(x = z, fill = z_quant)) +
  geom_histogram(alpha = 0.50) +
  facet_wrap(facets = ~ x_quant)

```


```{r}

lm_01 %>% effect(mod = .,
                 term = "x:cat:z",
                 x.var = "x",
                 xlevels = list(cat = c(-1, 1),
                                z = quantile(example_data$z))) %>% 
  as_tibble() %>% 
  mutate(cat = cat %>% factor(),
         z = z %>% round(2) %>% factor()) %>% 
  ggplot(mapping = aes(x = x, y = fit, 
                       
                       
                       
                       
                       
                       
                       
                       
                       
                       
                       cat)) +
  facet_wrap(facets = ~ z) +
  geom_point(data = example_data %>%
               dplyr::select(-z), aes(x = x, y = y, color = factor(cat))) +
  geom_smooth(method = "lm") +
  geom_ribbon(alpha = 0.25, aes(ymin = lower, ymax = upper, fill = cat, color = NULL))

```

## test conditional regression slopes
> get quantile values for z; save them

```{r}

zqs <- quantile(example_data$z)

```

#### 0% quantile of z, cat = -1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, -1, zqs[1], 0, -1 * zqs[1]))) %>% 
  summary(test = adjusted("none"))

```

#### 0% quantile of z, cat = 1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, 1, zqs[1], 0, 1 * zqs[1]))) %>% 
  summary(test = adjusted("none"))

```

#### 25% quantile of z, cat = -1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, -1, zqs[2], 0, -1 * zqs[2]))) %>% 
  summary(test = adjusted("none"))

```

#### 25% quantile of z, cat = 1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, 1, zqs[2], 0, 1 * zqs[2]))) %>% 
  summary(test = adjusted("none"))

```

#### 50% quantile of z, cat = -1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, -1, zqs[3], 0, -1 * zqs[3]))) %>% 
  summary(test = adjusted("none"))

```

#### 50% quantile of z, cat = 1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, 1, zqs[3], 0, 1 * zqs[3]))) %>% 
  summary(test = adjusted("none"))

```

#### 75% quantile of z, cat = -1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, -1, zqs[4], 0, -1 * zqs[4]))) %>% 
  summary(test = adjusted("none"))

```

#### 75% quantile of z, cat = 1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, 1, zqs[4], 0, 1 * zqs[4]))) %>% 
  summary(test = adjusted("none"))

```

#### 100% quantile of z, cat = -1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, -1, zqs[5], 0, -1 * zqs[5]))) %>% 
  summary(test = adjusted("none"))

```

#### 100% quantile of z, cat = 1

```{r}

lm_01 %>%
  glht(linfct = rbind(c(0, 1, 0, 0, 1, zqs[5], 0, 1 * zqs[5]))) %>% 
  summary(test = adjusted("none"))

```












