---
title: "Testing parameters in structural equation models"
author: "Nicholas Michalak"
date: "6/12/2017"
output: html_document
---

# packages

```{r}

# packages wanted for script
want_packages <- c("lavaan", "tidyverse")

# wanted packages in list of installed packages
have_packages <- want_packages %in% rownames(installed.packages())

# if any packages aren't in the installed list
if(any(!have_packages) == TRUE)

# install them
install.packages(want_packages[!have_packages])

# load libraries
lapply(want_packages, library, character.only = TRUE)
```

# covariance matrix

```{r}

# The classic Wheaton et. al. (1977) model
# panel data on he stability of alienation
wheaton_cov <- "11.834,
                6.947, 9.364,
                6.819, 5.091, 12.532,
                4.783, 5.028, 7.495, 9.986"

wheaton_names <- paste0("V", 1:4)

# convert to a full symmetric covariance matrix with names
wheaton_cov <- getCov(wheaton_cov,
                      lower = TRUE,
                      diagonal = TRUE,
                      names = wheaton_names)
```

>  In this simple model there are two latent variables, each with two indicators; there are also equality constraints on the error variances, *V*(E1) = *V*(E3) = \(\phi_1\) and *V*(E2) = *V*(E4) = \(\phi_2\), and the loadings. The model and data are taken from the EQS manual (Bentler, 1995). The critical parameter for our illustration is the covariance between the two latent variables (\(\phi_{12}\)). Restrictions are needed in order to set the scale of the latent variables. The analyst may fix the variance of each of the latent variables to a constant or may fix indicator coefficients to a constant (by convention, this constant is chosen to be 1; e.g., Bollen, 1989). We refer to such restrictions as *model identifications*. Figures 1 and 2 illustrate the model under two different identifications, or scale restrictions. The model in Figure 1 forces one indicator loading to 1 within each latent variable and treats the latent variances as parameters to estimate, whereas the model in Figure 2 forces the latent variances to 1 and treats the indicator loadings as parameters. Except for this difference in identification, the two models are equivalent, and both models produce identical fits to the observed covariance matrix.

```{r}
# Figure 1 model
# V1 and V3 estimates should be 1
# V2 and V4 estimates should be free
# cov for V1 and V3 should be equal
# cov for V2 and V4 should be equal
model_1 <- "# latent 1
              latent_1 =~ V1 + l1*V2

            # latent 2
              latent_2 =~ V3 + l1*V4
          
            # covariances
              V1 ~~ a*V1
              V2 ~~ b*V2
              V3 ~~ a*V3
              V4 ~~ b*V4
              V3 ~~ co*V1
              V4 ~~ co*V2
              latent_1 ~~ lat_var*latent_1
              latent_2 ~~ lat_var*latent_2"

# fit sem model
fit_1 <- sem(model_1,
             sample.cov = wheaton_cov,
             sample.nobs = 932)

# fit measures
fitmeasures(fit_1)
```

```{r}
# Figure 2 model
# free estimate all exogenous
# latent 1 and latent 2 should be 1
# exogenous var for V1 and V3 should be equal
# exogenous var for V2 and V4 should be equal
model_2 <- "# latent 1
              latent_1 =~ NA*V1 + NA*V2

            # latent 2
              latent_2 =~ NA*V3 + NA*V4
          
            # covariances
              V1 ~~ a*V1
              V2 ~~ b*V2
              V3 ~~ a*V3
              V4 ~~ b*V4
              V3 ~~ co*V1
              V4 ~~ co*V2
              latent_1 ~~ 1*latent_1
              latent_2 ~~ 1*latent_2"

# fit sem model
fit_2 <- sem(model_2,
             sample.cov = wheaton_cov,
             sample.nobs = 932)

# fit measures
fitmeasures(fit_2)
```
